Studying public transit API query logs\\ to get an indication of travel flows

### 2016-01-08 paper.tex

\emph{Flow analysis} is a topic of theoretical interest and practical importance in various disciplines. 
Flow analysis is conventionally conducted to study spatial dynamics and identify routine patterns in the movement of people.
For instance, interest in modelling traffic flows emerged from the strain placed on urban transportation systems during peak hours~\cite{roth,ferreira}.
Likewise, insight into routine travel patterns is crucial for the conceptualisation of functional urban areas~\cite{servillo2013town}, urban hierarchies~\cite{christaller1966central} and other territorial structures.

@@

We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
The other features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
Each row within this dataset contains:
\begin{enumerate}
\item A timestamp of when the request was executed,
\item The request path between two stations and their associated geographic coordinates,
\item The user-agent string cfr. RFC2616,
\item The operating system of the user-agent and
\item A flag stating if the user-agent is a bot.
\end{enumerate}

@@

In this chapter, we report on the results of studying visualizations with experts of \emph{Trein Tram Bus}\footnote{\url{http://treintrambus.be}}, a not for profit passenger federation.
%Isn't this sufficient ↑? \todo{Trein Tram Bus needs to be introduced in the introduction}

@@

The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively. 
Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.
%We can observe that Brussels, Antwerp, Ghent and Leuven are the dominant destinations.
%Turning our attention to weekday travel, we can observe that Antwerp, Brussels and Ghent are the dominant destinations.
%The reverse occurs in the evening when trips are made towards provincial stations.
Brussels City stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity (see \cref{fig:mornings}a).
On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
With exception to Liège city (see \cref{fig:mornings}e), queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Ghent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d). 
These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels city as a major centre.
The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.

>>>>

\emph{Flow analysis} is a topic of theoretical interest and practical importance in various disciplines. 
Flow analysis is conventionally conducted to study spatial dynamics and identify routine patterns in the movement of people.
For instance, interest in modelling traffic flows emerged from the strain placed on urban transportation systems during peak hours~\cite{roth,ferreira}.
Likewise, insight into routine travel patterns is crucial for the conceptualization of functional urban areas~\cite{servillo2013town}, urban hierarchies~\cite{christaller1966central} and other territorial structures.

@@

We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
The other features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
We made a copy of this dataset available in CSV format can be downloaded at \url{https://github.com/pietercolpaert/irail-route-planning-query-logs}.
%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
Each row within this dataset contains:
\begin{enumerate}
\item A timestamp of when the request was executed,
\item The request path between two stations and their associated geographic coordinates,
\item The user-agent string cfr. RFC2616,
\item The operating system of the user-agent and
\item A flag stating if the user-agent is a bot.
\end{enumerate}

@@

In this chapter, we report on the results of studying visualizations with experts of \emph{Trein Tram Bus}\footnote{\url{http://treintrambus.be}}, a not for profit passenger federation.

@@

The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively\footnote{An interactive version of this visualization can be viewed at \url{http://rxd.architectuur.kuleuven.be/irail/weekdays.html} and \url{http://rxd.architectuur.kuleuven.be/irail/weekends.html}}. 
Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.
%We can observe that Brussels, Antwerp, Ghent and Leuven are the dominant destinations.
%Turning our attention to weekday travel, we can observe that Antwerp, Brussels and Ghent are the dominant destinations.
%The reverse occurs in the evening when trips are made towards provincial stations.
Brussels City stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity (see \cref{fig:mornings}a).
On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
With exception to Liège city (see \cref{fig:mornings}e), queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Ghent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d). 
These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels city as a major centre.
The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.
